Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-06-13 paper-new/technical-background.tex
[paper] Update background on Ticked Cubical Type Theory

\section{Background on Guarded Domain Theory}\label{sec:technical-background}

@@

There is a \emph{guarded fixpoint} operator

\[
  \fix : \forall T, (\later T \to T) \to T.
\]

That is, to construct a term of type $T$, it suffices to assume that we have access to
such a term ``later'' and use that to help us build a term ``now''.
This operator satisfies the axiom that $\fix f = f (\nxt (\fix f))$.
In particular, this axiom applies to propositions $P : \texttt{Prop}$; proving
a statement in this manner is known as $\lob$-induction.

@@

The rules for tick abstraction and application are similar to those of ordinary
$\Pi$ types. In particular, if in a context $\Gamma, t : \tick\, k$ we have that $M$
has type $A$, then in context $\Gamma$ we have $\lambda t.M$ has type $\later^k
A$. % TODO dependent version?
%
Conversely, if we have a term $M$ of type $\later^k A$, and we have available in
the context a tick $t' : \tick\, k$, then we can apply the tick to $M$ to get a
term $M[t'] : A[t'/t]$. % TODO restriction on tick application
%
For the sake of brevity, we will also write tick application as $M_t$.

>>>>

\section{Background on Synthetic Guarded Domain Theory}\label{sec:technical-background}

@@

There is a \emph{guarded fixpoint} operator
%
\[
  \fix : \forall T, (\later T \to T) \to T.
\]
%
That is, to construct a term of type $T$, it suffices to assume that we have access to
such a term ``later'' and use that to help us build a term ``now''.
This operator satisfies the axiom that $\fix f = f (\nxt (\fix f))$.
In particular, this axiom applies to propositions $P : \texttt{Prop}$; proving
a statement in this manner is known as $\lob$-induction.

@@

% TODO include a figure with some of the rules for ticks

The rules for tick abstraction and application are similar to those of ordinary
$\Pi$ types. A context now consists of ordinary variables $x : A$ as well as
tick variables $t : \tick\, k$. The presence of the tick variable $t$ in context
$\Gamma, (t : \tick\, k), \Gamma'$ intuitively means that the values of the
variables in $\Gamma$ arrive ``first'', then one time step occurs on $k$, and
then the values of the variables in $\Gamma'$ arrive.

The abstraction rule for ticks states that if in context $\Gamma, t : \tick\, k$
the term $M$ has type $A$, then in context $\Gamma$ the term $\lambda t.M$ has
type $\later^k_t A$. % TODO dependent version?
%
Conversely, if we have a term $M$ of type $\later^k A$, and we have available in
the context a tick $t' : \tick\, k$, then we can apply the tick to $M$ to get a
term $M[t'] : A[t'/t]$. However, there is an important restriction on when we
are allowed to apply ticks. In order to apply $M$ to tick $t$, $M$ must be
well-typed in the prefix of the context occurring before the tick $t$. That is,
all variables mentioned in $M$ must be available before $t$. This ensures that
we cannot, for example, define a term of type $\later \laterhs A \to\, \laterhs
A$ via repeated tick application.
% TODO restriction on tick application
%
For the sake of brevity, we will also write tick application as $M_t$.
